evidence of adaption-independent time-warp spike timing in cortical networks

### 2020-07-29 manuscript.tex
Jolly goood

\title{evidence of stationary and non-stationary clustered time-warped spike timing in cortical networks}

@@

Responses following sensory stimulation can be analysed to explore principles of neural coding. Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. A multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space (Fig. 1a). 
% which approximates an underlying stimulus-response distribution.

@@

Information theoretic approaches aim to understand 
If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Characterising the distribution from which a stimulus response is `drawn' from on a single trial is complicated by correlative structure, neural adaptation and heterogeneity of sample response distribution structure.

Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of noise shared between the responses of neurons. These are referred to as `noise correlations' and are the focus of this manuscript, rather than `signal correlations' which refer to tuning similarity between neurons over multiple stimuli.

Information theoretic approaches classically consider neural encoding with respect to fixed-state encoding and decoding. These approaches often do not consider that correlative structure (i.e. covariance of multi-neuron responses over trials of a single stimulus) could be caused by a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.

Firstly, correlative structure between the responses of spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) in a single stimulus response distribution, opens the possibility of a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial. Response samples on a single trial, when the continuous state has a certain value, would then come from a state-conditioned stimulus-response distribution. This would be a smaller region of stimulus-response space than the region from which all samples come (Fig. 1c). Therefore, if downstream neurons could respond in a state-dependent manner (contrary to common neural coding assumptions \cite{moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding.

Characterisation of response distribution structure is also complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}. That is, the response distributions are \textit{non-stationary} and change as a function of trial index for some neurons. Adaptation could therefore lead to correlative structure in single stimulus response distributions. Adaptation must therefore be controlled for when searching for correlative structure indicative of a possible underlying state.

The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}. The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering controlled study of the neural representation of atomic entities. The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for atomic representation of information. Representation by single spike patterns is supported in other sensory modalities, such as vision, where a recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.

First spike times with millisecond order precision have been observed relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}. Relative differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} have been shown to encode information when trial averaged \cite{gollisch2008rapid} and do not require reference to an onset signal.
Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come on a single trial (Fig. 1d; \cite{reyes2014laminar}).
% as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units. That is, for repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. The reader is strongly recommended to view these plots in the top left of Supplementary Vid. 1. Simple observation of these plots (ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.

@@

Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1).
A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). The observed correlations with `non-unitary gradient' may suggest that correlations are due to an internal process, for example, a shared cortical state underlying state-dependent processing. Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.

A novel algorithm extracted the observed clusters and found that they were correlated above chance levels. To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity. The cluster spike times of both units were stationary for X\% of correlated clusters. X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying state co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig. 1d).

For the remaining clusters, with at least one unit with non-stationary cluster spike times, time series analysis was applied to remove the neural adaptive trends of the non-stationary unit/units. X\% of the detrended clusters were correlated with non-unitary gradient, suggestive of an underlying state.

Application of state analysis to the stationary and detrended clusters provided an estimate of possible state-dependent response distributions. This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms and controls for neural adaptive trends. It also provides a precise understanding of stimulus responses and correlative variability, which could be interpreted classically (i.e. independently of state [Vicente Refs]).

The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature. It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. Testing the state-dependent decoding performance of the response properties uncovered in this paper would require explicit knowledge (or an estimate) of the underlying trial-to-trial state so is left for future studies.
Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.

The characterised response distribution structure is referred to as `stationary and non-stationary clustered time-warped spike timing'. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner.

The results suggest that consideration of clustering, state-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics. In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.

>>>>

\title{evidence of adaption-independent time-warp spike timing in cortical networks}

@@

Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
A multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution (Fig. 1a). Estimating the distribution from which a stimulus response is `drawn' from on a single trial is complicated by correlative structure, neural adaptation and heterogeneity of sample response distribution structure.
% which approximates an underlying stimulus-response distribution.

@@

% Information theoretic approaches aim to understand 
If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural} using approaches from information theory.

Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons for repeated presentations of a single stimulus \cite{averbeck2006neural}. This paper considers the precise nature of noise correlations for single stimuli. Noise correlations differ from `signal correlations' which concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}). Information theory approaches classically consider neural coding and noise correlations with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
% (and so the stimulus-dependency of noise correlations is left to future studies).

% Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of trial-to-trial noise shared between the responses of multiple neurons. These are classically referred to as `noise correlations' \cite{averbeck2006neural}, rather than   
It is important, however, to consider that noise correlations between spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) could be caused by an underlying \textit{continuous} single or low-dimensional latent state covarying neural responses from trial-to-trial. A response on a single trial, when the continuous latent state has a certain value, would then come from a \textit{state-conditioned} stimulus-response distribution. This would be a smaller region of stimulus-response space than the region from which all samples are drawn from over multiple presentations of a single stimulus (Fig. 1c). Therefore, if downstream neurons could respond in a state-dependent manner (contrary to classical neural coding assumptions \cite{averbeck2006neural, moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding.

Characterising response correlations and stimulus-response distributions at a point in time is complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. Processes of adaptation, which we define as different to possible \textit{non-adaptive} underlying latent states, could also lead to observed correlative structure between neuronal responses (Fig. X). Adaptation must therefore be controlled for when characterising correlative structure, which could be indicative of \textit{non-adaptive} underlying latent states.

The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}. The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents. The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for the representation of atomic stimuli. Representation by single spike patterns is supported in other sensory modalities, such as vision, where a recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}. Adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.

[Up to here]

First spike times with millisecond order precision have been observed relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}. Relative spike time differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} do not require reference to an onset signal and have been shown to encode information when trial averaged \cite{gollisch2008rapid}.
Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations of a single stimulus come on a single trial (Fig. 1d; \cite{reyes2014laminar}).
% as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units. That is, for repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. The reader is strongly recommended to view these plots in the top left of Supplementary Vid. 1. Simple observation (ignoring lines, colours and ellipses at this stage) of these single stimulus first spike pair response distributions shows the presence of isolated clusters, which often appear positively correlated. A novel algorithm extracted the observed clusters and found that they were positively correlated above chance levels, showing that first spike pairs within isolated regions of response distribution space are positively correlated above chance-levels.

@@

Further observation of the response distributions, shows that the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1). A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). The observed correlations, instead have `non-unitary gradients' (different from $45^{\circ}$), which may be indicative that correlations are due to an internal process, for example, a shared latent state underlying state-dependent processing. Moreover, non-unitary correlation angles would suggest that the distribution of decodable relative spike time differences for the pair of neurons also varies with respect to the underlying latent state. Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.

To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity. The cluster spike times of both units were stationary for X\% of correlated clusters. X\% of these stationary correlated clusters had angles of correlation which were significantly different from $45^{\circ}$ (non-unitary), suggestive of an internal underlying latent state co-varying first spike times (and relative differences) from trial-to-trial (Fig. 1d). 
For the remaining clusters, with at least one unit with non-stationary cluster spike times, time series analysis was applied to remove the neural adaptive trends of the non-stationary unit/units. X\% of the detrended clusters were correlated with non-unitary gradient, suggestive of an underlying state. The analyses also provides a useful characterisation of spike-timing correlative variability for single stimuli, which could be useful to understanding how different stimuli are encoded under the classical neural coding interpretation (i.e. independently of a shared state).

Application of factor analysis to the stationary and detrended clusters provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences). This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms. Correlated clusters appeared for pairs of neurons which were spatially separated over multiple barrels and cortical layers. The non-unitary positive correlations between spatially distanced neurons is tantalisingly suggestive of a single shared lantent state, that could also modulate downstream neuron and allow them to respond in a state-dependent manner, with the benefits of high-capacity state-dependent encoding. The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed.

Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came. [I have begun developing several techniques to address this but I think it may prove difficult to show anything strongly conclusive either way. I can give more details of this once we are agreed on the overall objectives. We may be able to address this issue, but it may turnout to require future experiments designed to get a higher cluster count. A bit of help attacking this problem and working on factor analysis in 3+ dimensions would also be helpful.]

The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature. 
% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
Testing the state-dependent decoding performance of the response properties uncovered in this paper is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts).
Subthreshold membrane potentials would be a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}. The LFP may offer a useful proxy for the subthreshold membrane potential which could predict the correlations uncovered in this paper but it may prove useful to determine the dimensionality of the states underlying the observed correlations first.

The characterised response distribution structure is referred to as `adaption-independent clustered time-warped spike timing' and describes that first spike time clusters for pairs of neurons are correlated with non-unitary gradient, such that the relative spike timing encoding is `warped' from trial-to-trial by some `non-adaptive' continuous latent state. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner.

The results suggest that consideration of clustering, correlative structure and adaption, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics. In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.
